Cell type assignments for SCPCP000015

### 2025-02-03 analyses/cell-type-ewings/exploratory_analysis/08-merged-celltypes.Rmd
notebook for looking at merged results

---
title: "Template notebook for validating cell type assignments for an individual library in SCPCP000015"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: "hide"
params:
  sample_id: "SCPCS000490"
  library_id: "SCPCL000822" 
  cluster_nn: 20
  cluster_res: 0.5
---

This notebook is meant to be a guide for compiling "final" cell type annotations for an individual library in `SCPCP000015`. 
Results from `aucell-singler-annotation.sh`, `evaulate-clusters.sh`, and `run-aucell-ews-signatures.sh` are all combined and used to assign cell type annotations on a case by case basis.

Instructions for using this guide:

1. Ensure that you have a local copy of the results from `aucell-singler-annotation.sh`, `evaluate-clusters.sh` and `run-aucell-ews-signatures.sh` saved to `results`. 
2. Copy the contents of this notebook to a new notebook titled `<library_id>_celltype-exploration.Rmd` and save in `exploratory_analysis/final_annotation_notebooks`. 
3. Update the `title` in the `yaml` section and replace the `sample_id` and `library_id` with the correct IDs in the `params` list. 
4. Optionally, you may choose to update the choices for clustering based on the results from `evaluate-clusters.sh`. 
All clusters used will be calculated with the Leiden algorithm and the modularity objective function. 
To modify the nearest neighbors (default: 20) and resolution (default: 0.5) chosen use the `cluster_nn` and `cluster_res` params. 
5. Run through the notebook and update any sections of the notebook marked with the `{.manual-exploration}` tag. 
6. Render the completed notebook which will produce the rendered `html` file and a TSV with cell type annotations for that library.

@@

# The current data directory, found within the repository base directory
data_dir <- file.path(repository_base, "data", "current", "SCPCP000015")

@@

```{r}
# path to sce 
sce_file <- file.path(data_dir, params$sample_id, glue::glue("{params$library_id}_processed.rds"))

@@

singler_results_dir <- file.path(workflow_results_dir, "aucell_singler_annotation", params$sample_id)
singler_results_file <- file.path(singler_results_dir,
                                  glue::glue("{params$library_id}_singler-classifications.tsv"))

cluster_results_dir <- file.path(workflow_results_dir, "clustering", params$sample_id)
cluster_results_file <- file.path(cluster_results_dir,
                                  glue::glue("{params$library_id}_cluster-results.tsv"))

aucell_results_dir <- file.path(workflow_results_dir, "aucell-ews-signatures", params$sample_id)
aucell_results_file <- file.path(aucell_results_dir,
                                 glue::glue("{params$library_id}_auc-ews-gene-signatures.tsv"))

@@

```{r}
# output file to save final annotations 
results_dir <- file.path(module_base, "results", "final-annotations")
output_file <- file.path(results_dir, glue::glue("{params$library_id}_celltype-annotations.tsv"))
```

@@

```{r}
stopifnot(
  "sce file does not exist" = file.exists(sce_file),
  "singler results file does not exist" = file.exists(singler_results_file),
  "cluster results file does not exist" = file.exists(cluster_results_file),
  "aucell results file does not exist" = file.exists(aucell_results_file)
)
```

@@

# read in workflow results
singler_df <- readr::read_tsv(singler_results_file)
cluster_df <- readr::read_tsv(cluster_results_file)
aucell_df <- readr::read_tsv(aucell_results_file)

@@

```{r}
all_results_df <- prep_results(
  sce, 
  singler_df, 
  cluster_df, 
  aucell_df,
  cluster_nn = params$cluster_nn,
  cluster_res = params$cluster_res
  )

@@

all_info_df <- all_results_df |> 
  dplyr::left_join(gene_exp_df, by = "barcodes") |> 
  dplyr::arrange(cluster)
```

@@

# reformat auc data for density plots and UMAPs showing AUC values 
auc_df <- all_info_df |> 
  tidyr::pivot_longer(starts_with("auc_"), names_to = "geneset", values_to = "auc_value") |> 
  dplyr::mutate(
    geneset = stringr::str_remove(geneset, "auc_")
  ) |> 
  dplyr::select(barcodes, UMAP1, UMAP2, geneset, auc_value, cluster) |> 
  dplyr::left_join(auc_threshold_df, by = c("barcodes", "geneset")) |> 
  dplyr::mutate(
    in_geneset =  auc_value > threshold
  )
```

@@

```{r}
annotated_exp_heatmap(
  all_info_df, 
  exp_columns = auc_columns, 
  cell_type_column = "singler_lumped", 
  cluster_column = "cluster", 
  legend_title = "AUC"
)
```

@@

```{r, fig.height=8}
ggplot(exp_df, aes(x = mean, color = tumor_cell_classification, fill = tumor_cell_classification)) +
  geom_density(alpha = 0.5, bw = 0.2) +
  facet_wrap(vars(geneset)) +
  theme(
    aspect.ratio = 1,
    strip.background = element_rect(fill = "transparent", linewidth = 0.5),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5)
  )
```

@@

```{r}
annotated_exp_heatmap(
  all_info_df, 
  exp_columns = mean_exp_columns, 
  cell_type_column = "singler_lumped", 
  cluster_column = "cluster", 
  legend_title = "Mean\ngene expression"
)
```

>>>>

---
title: "Cell type assignments for SCPCP000015"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: "hide"
---

@@

# The current data directory, found within the repository base directory
data_dir <- file.path(repository_base, "data", "current", "results", "merge-sce", "SCPCP000015")

@@

```{r}
# path to sce 
sce_file <- file.path(data_dir, "SCPCP000015_merged.rds")

@@

singler_results_dir <- file.path(workflow_results_dir, "aucell_singler_annotation")
singler_results_files <- list.files(singler_results_dir, pattern = "*singler-classifications.tsv", full.names = TRUE, recursive = TRUE)
library_ids <- stringr::str_remove(basename(singler_results_files), "_singler-classifications.tsv")

aucell_results_dir <- file.path(workflow_results_dir, "aucell-ews-signatures")
aucell_results_file <- file.path(aucell_results_dir, "SCPCP000015_auc-ews-gene-signatures.tsv")

@@

```{r}
# output file to save final annotations 
results_dir <- file.path(module_base, "results", "final-annotations")
output_file <- file.path(results_dir, glue::glue("SCPCP000015_celltype-annotations.tsv"))
```

@@

```{r}
stopifnot(
  "sce file does not exist" = file.exists(sce_file),
  "aucell results file does not exist" = file.exists(aucell_results_file)
)
```

@@

# read in workflow results
singler_df <- singler_results_files |> 
  purrr::set_names(library_ids) |>
  purrr::map(readr::read_tsv) |> 
  dplyr::bind_rows(.id = "library_id")
aucell_df <- readr::read_tsv(aucell_results_file) |> 
  tidyr::separate(barcodes, "-", into = c("library_id", "barcodes"))

@@

```{r}
all_results_df <- prep_results(
  sce, 
  singler_df = singler_df, 
  cluster_df = NULL, 
  aucell_df = aucell_df,
  cluster_nn = params$cluster_nn,
  cluster_res = params$cluster_res,
  join_columns = c("barcodes", "library_id")
  ) |>
  dplyr::mutate(barcodes = glue::glue("{library_id}-{barcodes}"))

@@

all_info_df <- all_results_df |> 
  dplyr::left_join(gene_exp_df, by = "barcodes")
```

@@

# reformat auc data for density plots and UMAPs showing AUC values 
auc_df <- all_info_df |> 
  tidyr::pivot_longer(starts_with("auc_"), names_to = "geneset", values_to = "auc_value") |> 
  dplyr::mutate(
    geneset = stringr::str_remove(geneset, "auc_")
  ) |> 
  dplyr::select(barcodes, UMAP1, UMAP2, geneset, auc_value) |> 
  dplyr::left_join(auc_threshold_df, by = c("barcodes", "geneset")) |> 
  dplyr::mutate(
    in_geneset =  auc_value > threshold
  )
```

@@

```{r}
all_info_df <- all_info_df |> 
  dplyr::arrange(singler_lumped)
head(all_info_df)

annotated_exp_heatmap(
  all_info_df, 
  exp_columns = auc_columns, 
  cell_type_column = "singler_lumped", 
  cluster_column = "singler_lumped", 
  legend_title = "AUC"
)
```

@@

```{r, fig.height=8}
ggplot(exp_df, aes(x = mean, color = tumor_cell_classification, fill = tumor_cell_classification)) +
  geom_density(alpha = 0.5, bw = 0.1) +
  facet_wrap(vars(geneset)) +
  theme(
    aspect.ratio = 1,
    strip.background = element_rect(fill = "transparent", linewidth = 0.5),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5)
  )
```

@@

```{r}
annotated_exp_heatmap(
  all_info_df, 
  exp_columns = mean_exp_columns, 
  cell_type_column = "singler_lumped", 
  cluster_column = "singler_lumped", 
  legend_title = "Mean\ngene expression"
)
```
